Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-04-25 doc/thesis/intro.tex
Writing

In epidemiology, a \defn{transmission tree} is a graphical representation of an
epidemic's progress through a population. Like phylogenies, transmission trees
have tips, nodes, edges, and branch lengths. However, rather than recording an
evolutionary process (speciation), they record an epidemiological process
(transmission). The tips of a transmission tree represent the removal of
infected hosts, while internal nodes correspond to transmissions from one host
to another.Transmission trees generally have branch lengths in units of
calendar time, with branching times indicating times of transmission. The root
of a transmission tree corresponds to the initially infected patient who
introduced the epidemic into the network, also known as the \defn{index case}.
The internal nodes may be labelled with the donor of the transmission pair, if
this is known. The tips of the tree, rather than being fixed at the present
day, are placed at the time at which the individual was removed from the
epidemic, such as by death, recovery, isolation, behaviour change, or
migration. Consequently, the transmission tree may not be ultrametric, but may
have tips located at varying distances from the root. Such trees are said to
have \defn{heterochronous} taxa~\autocite{drummond2003measurably}, in contrast
to the \defn{isochronous} taxa found in most phylogenies of macro-organisms. A
transmission tree is illustrated in \cref{fig:contactnet} (right).

@@

The basis of phylodynamics is the fact that, for RNA viruses, epidemiological
and evolutionary processes occur on similar time
scales~\autocite{drummond2003measurably}. In fact, these two processes
interact, such that it is possible to detect the influence of host epidemiology
on the evolutionary history of the virus as recorded in an \defn{inter-host
viral phylogeny}. Phylodynamic methods aim to detect and quantify the
signatures of epidemiological processes in these
phylogenies~\autocite{pybus2009evolutionary, volz2013viral}, which relate one
representative viral genotype from each host in an infected population. These
methods have been used to investigate parameters such as transmission rate,
recovery rate, and basic reproductive number~\autocite{pybus2009evolutionary,
volz2013viral}. The majority of phylodynamic studies attempt to infer the
parameters of an epidemiological model for which the likelihood of an observed
phylogeny can be calculated. Most often, this is some variation of the
birth-death~\autocite{kendall1948generalized, stadler2012estimating} or
coalescent~\autocite{kingman1982coalescent, volz2012complex} models. These
methods generally assume the viral phylogeny is known, and we assume the same
in this work (see \eg \autocite{nei2000molecular} for a review of phylogenetic
inference).

Due to the relationship between the aforementioned processes, there is a degree
of correspondence between viral phylogenies and transmission
trees~\autocite{ypma2013relating, kenah2015algorithms, kenah2016molecular}. In
particular, the transmission process is quite similar to \defn{allopatric
speciation}~\autocite{coyne2004speciation}, where genetic divergence follows
the geographic isolation of a sub-population of organisms. Thus, transmission,
which is represented as branching in the transmission tree, causes branching in
the viral phylogeny as well. Similarly, the removal of an individual from the
transmission tree causes the extinction of their viral lineage in the
phylogeny. Consequently, the topology of the viral phylogeny is sometimes used
as a proxy for the topology of the transmission
tree~\autocite{hall2015epidemic}. Modern likelihood-based methods of
phylogenetic reconstruction~\autocite[\eg][]{price2010fasttree,
stamatakis2014raxml} produce unrooted trees whose branch lengths measure
genetic distance in units of expected substitutions per site. On the other
hand, transmission trees are rooted, and have branches measuring calendar
time~\autocite{pybus2009evolutionary}. Therefore, estimating a transmission
tree from a viral phylogeny requires the phylogeny to be rooted and
time-scaled. Methods for performing this process include root-to-tip
regression~\autocite{shankarappa1999consistent, korber2000timing,
drummond2003inference}, which we apply in this work, and least-square
dating~\autocite{to2015fast}. Alternatively, the tree may be rooted separately
with an outgroup~\autocite{li1988rates} before time-scaling.

A caveat of estimating transmission trees in this manner is that the
correspondence between the topologies of the viral phylogeny and transmission
tree is not necessarily exact~\autocite{ypma2013relating, romero2014timing}.
Due to intra-host diversity, the viral strain which is transmitted may have
split from another lineage within the donor long before the transmission event
occurred. Hence, the branching point in the viral phylogeny may be much earlier
than that in the transmission tree. Another possibility is that one host
transmitted to two or more recipients, but the lineages they each received
originated within the donor host in a different order than that in which the
transmissions occurred.  In this case, the topology of the transmission tree
and the viral phylogeny will be mismatched. In practice, this discordance has
not proven an insurmountable problem: for example,
\textcite{leitner1996accurate, paraskevis2004phylogenetic} were able to
accurately recover a known transmission tree using a viral phylogeny. The
problem of accurately estimating transmission trees is an ongoing area of
research~\autocite{cottam2008integrating, jombart2011reconstructing,
ypma2012unravelling, morelli2012bayesian, didelot2014bayesian}.

@@

\textcite{poon2013mapping} developed an alternative pairwise measure which
applies the concept of a \defn{kernel function} to phylogenies. Kernel
functions, originally developed for \glspl{SVM}~\autocite{burges1998tutorial}, 
compare objects in a space $\X$ by mapping them into a feature space $\F$ of
high or infinite dimension via a function $\varphi$. The similarity between the
objects is defined as
\[  
  K(x, x') = \langle \varphi(x), \varphi(x')\rangle,
\]
that is, the inner product of the objects' representations in the feature
space. Computing $\varphi(x)$ may be computationally prohibitive due to the
dimension of $\F$. The utility of a kernel function $K$ is that it is
constructed in such a way that it can compute the inner product without
explicitly computing $\varphi(x)$. The kernel function developed in
\autocite{poon2013mapping} will henceforth be referred to as the \defn{tree
kernel}. The feature space into which trees are mapped is the space of all
possible \defn{subset trees}, which are subtrees that do not necessarily extend
all the way to the tips. The subset-tree kernel was originally developed for
comparing parse trees in natural language processing~\autocite{collins2002new}
and did not incorporate branch length information. The version developed by
\textcite{poon2013mapping} includes a radial basis function to compare the
differences in branch lengths, thus incorporating both the trees' topologies
and their branch lengths in a single similarity score.

@@

Edges in a contact networks may be \defn{directed}, representing one-way
transmission risk, or \defn{undirected}, representing symmetric transmission
risk. For example, a network for an airborne epidemic would use undirected
edges, because the same physical proximity is required for a host to infect or
to become infected. However, an infection which may be spread through
blood-to-blood contact through transfusions transfusions would use directed
edges, since the donor has no chance of transmitting to the recipient. Directed
edges are also useful when the transmission risk is not equal between the
hosts, such as with HIV transmission, where acting as the receptive partner
carries a higher risk of infection than acting as the insertive partner. In
this case, a contact could be represented by two directed edges, one in each
direction between the two hosts, with the edges annotated by what kind of risk
they imply~\autocite{wasserman1994social}. An undirected contact network is
equivalent to a directed network where each contact is represented by two
symmetric directed edges. The \defn{degree} of a node in the network is how
many contacts it has. In directed networks, we may make the distinction between
\defn{out-degree} and \defn{in-degree}, which count respectively the number
incoming and outgoing edges. The \defn{degree distribution} of a network
denotes the probability that a node has any given number of links. The set of
edges attached to a node are referred to as its \defn{incident} edges.

Epidemiological models most often assume some form of contact homogeneity. The
simplest models, such as the \gls{SIR} model, assume a completely homogeneously
mixed population, where every pair of contacts is equally likely. More
sophisticated models partition the population into groups with different
contact rates between and among each group. However, these models still assume
that every possible contact between a member of group $i$ and a member of group
$j$ is equally likely. This assumption is clearly unrealistic for the majority
of human communities, and can lead to significant errors in predicted epidemic
trajectories when there is substantial heterogeneity
present~\autocite{bansal2007individual, volz2007susceptible,
rolls2015simulation}. Contact networks provide a way to relax this assumption
by representing individuals and their contacts explicitly. It is important to
note that, although panmixia is an unrealistic modelling assumption, it has not
proven a substantial hurdle to epidemic modelling in
practice~\autocite{anderson1992infectious}. Using this assumption, researchers
have been able to derive estimates of the transmission rate and the basic
reproductive number of various outbreaks, which have agreed with values
obtained by on-the-ground data collection. Therefore, if one is interested only
in these population-level variables, the additional complexity of contact
network models may not be warranted. Rather, these models are most useful when
we are interested in properties of the network itself, such as centrality,
structural balance, and transitivity~\autocite{wasserman1994social}.

From a public health perspective, knowledge of contact networks has the
potential to be extremely useful. On a population level, network structure can
dramatically affect the speed and pattern of epidemic
spread~\autocite[\eg][]{barthelemy2005dynamical, volz2008sir}. For example,
epidemics are expected to spread more rapidly in networks having the ``small
world'' property, where the average path length between two nodes in the
network is relatively low~\autocite{watts1998collective}. Some sexually
transmitted infections would not be expected to survive in a homogeneously
mixed population, but their long-term persistence can be explained by contact
heterogeneity~\autocite{anderson1992infectious, pastor2001epidemic}. Hence, the
contact network can provide an idea of what to expect as an epidemic unfolds.
In terms of actionable information, vaccination strategies which would
eradicate an epidemic in a random network might not work if the network is
scale-free~\autocite[][see \cref{subsec:pa}]{keeling2005networks}. On a local
level, contact networks can be informative about the groups or individuals who
are at highest risk of acquiring or transmitting infection, and would therefore
benefit most from public health interventions~\autocite{wang2015targeting,
little2014using}.

Contact networks are a challenging type of data to collect, requiring extensive
epidemiological investigation in the form of contact
tracing~\autocite{morris1993epidemiology, welch2011statistical,
keeling2005networks, eames2015six}. Therefore, it has been necessary to explore
less resource-intensive alternatives which still contain information about
population structure. For instance, it is possible to obtain limited
information about the contact network by individual interviews without contact
tracing. Variables which can be estimated in this fashion are referred to as
\defn{node-level} measures~\autocite{wasserman1994social}. One of the most
well-studied of these is the degree distribution, which can be estimated by
simply asking each person how many contacts they had in some interval of
time~\autocite{liljeros2001web, schneeberger2004scale, colgate1989risk}.

An alternative approach has been the analysis of other networks, which can be
estimated with phylogenetic methods from viral sequence data. Some work focuses
on the \defn{phylogenetic network}, in which two nodes are connected if the
genetic distance between their viral sequences is below some threshold.
Primarily, this work has focused on the detection of \defn{phylogenetic
clusters}, which are groups of individuals whose viral sequences are
significantly more similar to each other's than to the general population's.
The phylogenetic network is informative about ``hotspots'' of transmission and
can be used to identify demographic groups to whom targeted interventions are
likely to have the greatest effect~\autocite{poon2015impact}.  However, this
network may show little to no agreement with a contact data obtained through
epidemiological methods~\autocite{yirrell1998molecular, resik2007limitations,
robinson2013dynamics}, and therefore may be a poor proxy for the contact
network. Other studies~\autocite{brown2011transmission} have investigated the
\defn{transmission network}, which is the subgraph of the contact network
consisting of infected nodes and the edges which led to their
infections~\autocite{welch2011statistical} (\cref{fig:contactnet}, left). It is
possible to estimate the transmission network phylogenetically, although the
methods required for doing so are more sophisticated than for estimating the
phylogenetic network~\autocite{brown2011transmission}. These studies again
mostly focusing on clustering, and also on degree distributions.

@@

A number of studies have made progress in quantifying the relationship between
contact networks and transmission trees. \textcite{o2011contact} simulated
epidemics over networks with four types of degree distribution. They then
estimated the Bayesian skyride~\autocite{minin2008smooth} population size
trajectory in two ways: from the phylogeny, using \gls{MCMC}; and from the
incidence and prevalence trajectories, using the method developed by
\textcite{volz2009phylodynamics}. The concordance between the two skyrides, as
well as the relationship between the skyride and prevalence curve, was
qualitatively different for each degree distribution.
\textcite{leventhal2012inferring} investigated the relationship between
transmission tree imbalance and several epidemic parameters under four contact
network models, and found that these relationships varied considerably
depending on which model was being considered. \textcite{welch2011network}
simulated transmission trees over networks with varying degrees of community
structure. They found that transmission trees simulated under networks with low
clustering could not generally be distinguished from those simulated under
highly clustered networks, and concluded that contact network clusters do not
affect transmission tree shape. However, more recently,
\textcite{villandre2016assessment} investigated the correspondence between
contact network clusters and transmission tree clusters, and did find a
moderate correspondence between the two in some cases.

@@

We define a \defn{Markov kernel} as the continuous analogue of the transition
matrix in a finite-state Markov model.  For some spaces $X$ and $Y$, $K \maps X
\times Y \to [0, 1]$ such that
\begin{align}
    \label{eq:mk}
    \int_Y K(x, y) \d y = 1
\end{align}
for all $x \in X$. This is an ``operational'' definition of Markov kernel which
will be suitable for our purposes. A more rigorous definition can be found in
\eg \autocite{kallenberg2006foundations}. Note that a Markov kernels have
nothing to do with the kernel functions defined in \cref{subsec:treeshape},
other than sharing a name (the word ``kernel'' is ubiquitous in mathematics).

>>>>

In epidemiology, a \defn{transmission tree} is a graphical representation of an
epidemic's progress through a population. Like phylogenies, transmission trees
have tips, nodes, edges, and branch lengths. However, rather than recording an
evolutionary process (speciation), they record an epidemiological process
(transmission). The tips of a transmission tree represent the removal of
infected hosts, while internal nodes correspond to transmissions from one host
to another. Transmission trees generally have branch lengths in units of
calendar time, with branching times indicating times of transmission. The root
of a transmission tree corresponds to the initially infected patient who
introduced the epidemic into the network, also known as the \defn{index case}.
The internal nodes may be labelled with the donor of the transmission pair, if
this is known. The tips of the tree, rather than being fixed at the present
day, are placed at the time at which the individual was removed from the
epidemic, such as by death, recovery, isolation, behaviour change, or
migration. Consequently, the transmission tree may not be ultrametric, but may
have tips located at varying distances from the root. Such trees are said to
have \defn{heterochronous} taxa~\autocite{drummond2003measurably}, in contrast
to the \defn{isochronous} taxa found in most phylogenies of macro-organisms. A
transmission tree is illustrated in \cref{fig:contactnet} (right).

@@

The basis of phylodynamics is the fact that, for RNA viruses, epidemiological
and evolutionary processes occur on similar time
scales~\autocite{drummond2003measurably}. In fact, these two processes
interact, such that it is possible to detect the influence of host epidemiology
on the evolutionary history of the virus as recorded in an \defn{inter-host
viral phylogeny}. Phylodynamic methods aim to detect and quantify the
signatures of epidemiological processes in these
phylogenies~\autocite{pybus2009evolutionary, volz2013viral}, which relate one
representative viral genotype from each host in an infected population. These
methods have been used to investigate parameters such as transmission rate,
recovery rate, and basic reproductive number~\autocite{pybus2009evolutionary,
volz2013viral}. The majority of phylodynamic studies attempt to infer the
parameters of an epidemiological model for which the likelihood of an observed
phylogeny can be calculated. Most often, this is some variation of the
birth-death~\autocite{kendall1948generalized, stadler2012estimating} or
coalescent~\autocite{kingman1982coalescent, volz2012complex} models. These
methods either assume the viral phylogeny is known, as we do in this work, or
(more commonly) integrate over phylogenetic uncertainty using Bayesian methods.
Phylogenetic inference is a complex topic which we shall not discuss here; see
\eg \autocite{nei2000molecular} for a full review.

Due to the relationship between the aforementioned processes, there is a degree
of correspondence between viral phylogenies and transmission
trees~\autocite{leitner2002molecular, ypma2013relating, kenah2015algorithms,
kenah2016molecular}. In particular, the transmission process is quite similar
to \defn{allopatric speciation}~\autocite{coyne2004speciation}, where genetic
divergence follows the geographic isolation of a sub-population of organisms.
Thus, transmission, which is represented as branching in the transmission tree,
causes branching in the viral phylogeny as well. Similarly, the removal of an
individual from the transmission tree causes the extinction of their viral
lineage in the phylogeny. Consequently, the topology of the viral phylogeny is
sometimes used as a proxy for the topology of the transmission
tree~\autocite{hall2015epidemic}. Modern likelihood-based methods of
phylogenetic reconstruction~\autocite[\eg][]{price2010fasttree,
stamatakis2014raxml} produce unrooted trees whose branch lengths measure
genetic distance in units of expected substitutions per site. On the other
hand, transmission trees are rooted, and have branches measuring calendar
time~\autocite{pybus2009evolutionary}. Therefore, estimating a transmission
tree from a viral phylogeny requires the phylogeny to be rooted and
time-scaled. Methods for performing this process include root-to-tip
regression~\autocite{shankarappa1999consistent, korber2000timing,
drummond2003inference}, which we apply in this work, and least-square
dating~\autocite{to2015fast}. Alternatively, the tree may be rooted separately
with an outgroup~\autocite{li1988rates} before time-scaling.

A caveat of estimating transmission trees in this manner is that the
correspondence between the topologies of the viral phylogeny and transmission
tree is far from exact~\autocite{ypma2013relating, romero2014timing}.  Due to
intra-host diversity, the viral strain which is transmitted may have split from
another lineage within the donor long before the transmission event occurred.
Hence, the branching point in the viral phylogeny may be much earlier than that
in the transmission tree. Another possibility is that one host transmitted to
two or more recipients, but the lineages they each received originated within
the donor host in a different order than that in which the transmissions
occurred.  In this case, the topology of the transmission tree and the viral
phylogeny will be mismatched. In practice, this discordance has not proven an
insurmountable problem: for example, \textcite{leitner1996accurate,
paraskevis2004phylogenetic} were able to accurately recover a known
transmission tree using a viral phylogeny. The problem of accurately estimating
transmission trees is an ongoing area of
research~\autocite{cottam2008integrating, jombart2011reconstructing,
ypma2012unravelling, morelli2012bayesian, didelot2014bayesian}.

@@

\textcite{poon2013mapping} developed an alternative pairwise measure which
applies the concept of a \defn{kernel function} to phylogenies. Kernel
functions, originally developed for \glspl{SVM}~\autocite{burges1998tutorial}, 
compare objects in a space $\X$ by mapping them into a feature space $\F$ of
high or infinite dimension via a function $\varphi$. The similarity between the
objects is defined as
\[  
  K(x, x') = \langle \varphi(x), \varphi(x')\rangle,
\]
that is, the inner product of the objects' representations in the feature
space. Computing $\varphi(x)$ may be computationally prohibitive due to the
dimension of $\F$. The utility of a kernel function $K$ is that it is
constructed in such a way that it can compute the inner product without
explicitly computing $\varphi(x)$. The kernel function developed in
\autocite{poon2013mapping} will henceforth be referred to as the \defn{tree
kernel}. This kernel maps trees into the space of all possible possible
\defn{subset trees}, which are subtrees that do not necessarily extend all the
way to the tips. The subset-tree kernel was originally developed for comparing
parse trees in natural language processing~\autocite{collins2002new} and did
not incorporate branch length information. The version developed by
\textcite{poon2013mapping} includes a radial basis function to compare the
differences in branch lengths, thus incorporating both the trees' topologies
and their branch lengths in a single similarity score.

@@

Edges in a contact networks may be \defn{directed}, representing one-way
transmission risk, or \defn{undirected}, representing symmetric transmission
risk. For example, a network for an airborne epidemic would use undirected
edges, because the same physical proximity is required for a host to infect or
to become infected. However, an infection which may be spread through
blood-to-blood contact through transfusions transfusions would use directed
edges, since the donor has no chance of transmitting to the recipient. Directed
edges are also useful when the transmission risk is not equal between the
hosts, such as with \gls{HIV} transmission among \gls{MSM}, where the receptive
partner carries a higher risk of infection than the insertive
partner~\autocite{baggaley2010hiv}. In this case, a contact could be
represented by two directed edges, one in each direction between the two hosts,
with the edges annotated by what kind of risk they
imply~\autocite{wasserman1994social}. An undirected contact network is
equivalent to a directed network where each contact is represented by two
symmetric directed edges. The \defn{degree} of a node in the network is how
many contacts it has. In directed networks, we may make the distinction between
\defn{out-degree} and \defn{in-degree}, which count respectively the number
incoming and outgoing edges. The \defn{degree distribution} of a network
denotes the probability that a node has any given number of links. The set of
edges attached to a node are referred to as its \defn{incident} edges.

Epidemiological models most often assume some form of contact homogeneity. The
simplest models, such as the \gls{SIR} model~\autocite{anderson1992infectious},
assume a completely homogeneously mixed population, where every pair of
contacts is equally likely. More sophisticated models partition the population
into groups with different contact rates between and among each
group~\autocite{jacquez1988modeling}. However, these models still assume that
every possible contact between a member of group $i$ and a member of group $j$
is equally likely. This assumption is clearly unrealistic for the majority of
human communities, and can lead to significant errors in predicted epidemic
trajectories when there is substantial heterogeneity
present~\autocite{bansal2007individual, volz2007susceptible,
rolls2015simulation}. Contact networks provide a way to relax this assumption
by representing individuals and their contacts explicitly. It is important to
note that, although panmixia is an unrealistic modelling assumption, it has not
proven a substantial hurdle to epidemic modelling in
practice~\autocite{anderson1992infectious}. Using this assumption, researchers
have been able to derive estimates of the transmission rate and the basic
reproductive number of various outbreaks, which have agreed with values
obtained by on-the-ground data collection~\autocite{stadler2014insights}.
Therefore, if one is interested only in these population-level variables, the
additional complexity of contact network models may not be warranted. Rather,
these models are most useful when we are interested in properties of the
network itself, such as centrality, structural balance, and
transitivity~\autocite{wasserman1994social}.

From a public health perspective, knowledge of contact networks has the
potential to be extremely useful. On a population level, network structure can
dramatically affect the speed and pattern of epidemic
spread~\autocite[\eg][]{barthelemy2005dynamical, volz2008sir}. For example,
epidemics are expected to spread more rapidly in networks having the ``small
world'' property, where the average path length between two nodes in the
network is relatively low~\autocite{watts1998collective}. Some sexually
transmitted infections would not be expected to survive in a homogeneously
mixed population, but their long-term persistence can be explained by contact
heterogeneity~\autocite{anderson1992infectious, pastor2001epidemic}. Hence, the
contact network can provide an idea of what to expect as an epidemic unfolds.
In terms of actionable information, the efficacy of different vaccination
strategies may depend on the topology of the
network~\autocite{keeling2005networks,peng2013vaccination, ma2013importance,
rushmore2014network}. On a local level, contact networks can be informative
about the groups or individuals who are at highest risk of acquiring or
transmitting infection, and would therefore benefit most from public health
interventions~\autocite{wang2015targeting, little2014using}.

Contact networks are a challenging type of data to collect, requiring extensive
epidemiological investigation in the form of contact
tracing~\autocite{morris1993epidemiology, welch2011statistical,
keeling2005networks, eames2015six}. Therefore, it has been necessary to explore
less resource-intensive alternatives which still contain information about
population structure. For instance, it is possible to obtain limited
information about the contact network by individual interviews without contact
tracing. Variables which can be estimated in this fashion are referred to as
\defn{node-level} measures~\autocite{wasserman1994social}. One of the most
well-studied of these is the degree distribution mentioned above, which can 
theoretically be estimated by simply asking each person how many contacts they
had in some interval of time. However, the degree distributions often observed
in real-world sexual networks are heavy-tailed~\autocite{liljeros2001web,
schneeberger2004scale, colgate1989risk}, so dense or peer-driven sampling would
be needed to capture the high-degree nodes characterizing the tail of the
distribution.

An alternative approach has been the analysis of other types of network, which
can be directly estimated with phylogenetic methods from viral sequence data.
Some work focuses on the \defn{phylogenetic network}, in which two nodes are
connected if the genetic distance between their viral sequences is below some
threshold.  Primarily, this work has focused on the detection of
\defn{phylogenetic clusters}, which are groups of individuals whose viral
sequences are significantly more similar to each other's than to the general
population's.  The phylogenetic network is informative about ``hotspots'' of
transmission and can be used to identify demographic groups to whom targeted
interventions are likely to have the greatest effect~\autocite{poon2015impact}.
However, this network may show little to no agreement with a contact data
obtained through epidemiological methods~\autocite{yirrell1998molecular,
resik2007limitations, robinson2013dynamics}, and therefore may be a poor proxy
for the contact network. Other studies~\autocite{brown2011transmission} have
investigated the \defn{transmission network}, which is the subgraph of the
contact network consisting of infected nodes and the edges which led to their
infections~\autocite{welch2011statistical} (\cref{fig:contactnet}, left). It is
possible to estimate the transmission network phylogenetically, although the
methods required for doing so are more sophisticated than for estimating the
phylogenetic network~\autocite{brown2011transmission}. These studies again
mostly focusing on clustering, and also on degree distributions.

@@

A number of studies have made progress in quantifying the relationship between
contact networks and transmission trees. \textcite{o2011contact} simulated
epidemics over networks with four types of degree distribution. They then
estimated the Bayesian skyride~\autocite{minin2008smooth} population size
trajectory in two ways: from the phylogeny, using \gls{MCMC}; and from the
incidence and prevalence trajectories, using the method developed by
\textcite{volz2009phylodynamics}. The concordance between the two skyrides, as
well as the relationship between the skyride and prevalence curve, was
qualitatively different for each degree distribution.
\textcite{leventhal2012inferring} investigated the relationship between
transmission tree imbalance and several epidemic parameters under four contact
network models, and found that these relationships varied considerably
depending on which model was being considered. The authors also investigated a
real-world \gls{HIV} phylogeny and found a level of unbalancedeness
inconsistent with a randomly mixing population. \textcite{welch2011network}
simulated transmission trees over networks with varying degrees of community
structure. They found that transmission trees simulated under networks with low
clustering could not generally be distinguished from those simulated under
highly clustered networks, and concluded that contact network clusters do not
affect transmission tree shape. However, more recently,
\textcite{villandre2016assessment} investigated the correspondence between
contact network clusters and transmission tree clusters, and did find a
moderate correspondence between the two in some cases.

@@

We define a \defn{Markov kernel} as the continuous analogue of the transition
matrix in a finite-state Markov model.  For some spaces $X$ and $Y$, $K \maps X
\times Y \to [0, 1]$ such that
\begin{align}
    \label{eq:mk}
    \int_Y K(x, y) \d y = 1
\end{align}
for all $x \in X$. This is an ``operational'' definition of Markov kernel which
will be suitable for our purposes. A more rigorous definition can be found in
\eg~\autocite{kallenberg2006foundations}. Note that a Markov kernels have
nothing to do with the kernel functions defined in \cref{subsec:treeshape},
other than sharing a name (the word ``kernel'' is ubiquitous in mathematics).

### 2016-04-25 doc/thesis/objective.tex
Writing

The spread of a disease is most often modelled by assuming either a
homogeneously mixed population~\autocite{hamer1906milroy,
kermack1927contribution}, or a population divided into a small number of
homogeneously mixed groups~\autocite{rushton1955deterministic}. This
assumption, also called the ``law of mass action'' or \defn{panmixia}, implies
that any two individuals in the same compartment are equally likely to come
into contact causing transmission. Although this provides a reasonable
approximation in many cases~\autocite{anderson1992infectious}, the error
introduced by assuming a panmictic population can be substantial when
significant contact heterogeneity exists in the underlying population. Contact
network models provide an alternative to compartmental models which do not
require the assumption of panmixia. In addition to more accurate predictions,
the parameters of the networks themselves may be of interest from a public
health perspective. For example, certain vaccination strategies may be more or
less effective in curtailing an epidemic depending on the underlying network's
degree distribution~\autocite{peng2013vaccination}. Phylodynamic methods have
been used to fit many different types of model to phylogenetic
data~\autocite{pybus2009evolutionary,volz2013viral}, but as far as we know, no
methods have yet been developed to fit contact network models. The primary
objective of this work is to develop such a method.

@@

>>>>

The spread of a disease is most often modelled by assuming either a
homogeneously mixed population~\autocite{hamer1906milroy,
kermack1927contribution}, or a population divided into a small number of
homogeneously mixed groups~\autocite{rushton1955deterministic}. This
assumption, also called the ``law of mass action'' or \defn{panmixia}, implies
that any two individuals in the same compartment are equally likely to come
into contact causing transmission. Although this provides a reasonable
approximation in many cases~\autocite{anderson1992infectious}, the error
introduced by assuming a panmictic population can be substantial when
significant contact heterogeneity exists in the underlying population. Contact
network models provide an alternative to compartmental models which do not
require the assumption of panmixia. In addition to more accurate predictions,
the parameters of the networks themselves may be of interest from a public
health perspective. For example, certain vaccination strategies may be more or
less effective in curtailing an epidemic depending on the underlying network's
degree distribution~\autocite{peng2013vaccination, ma2013importance}.
Phylodynamic methods have been used to fit many different types of model to
phylogenetic data~\autocite{pybus2009evolutionary,volz2013viral}, but as far as
we know, no methods have yet been developed to fit contact network models. The
primary objective of this work is to develop such a method.

@@

\glsreset{ABC}
\glsreset{BA}
\glsreset{HIV}
\glsreset{ML}
\glsreset{SMC}
